**Title:** Lake Photic Zone Temperature across the Conterminous United States

### 2020-10-02 manuscript/manuscript.Rmd
this round of edits done...

---
title: "**Title:** Lake Photic Zone Temperature across the Conterminous United States"
author:
- ^*1*^ US Environmental Protection Agency, Office of Research and Development,  Atlantic
  Coastal Environmental Sciences Division (ACESD), Narragansett, RI 02882
- ^*2*^ ORISE, Narragansett, RI 02882
- '**Authors:** Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*2*^, Hollister. J. W. ^*1*^, Milstead,
  W. Bryan. ^*1*^'
- '^\*^  *corresponding author: kreakie.betty@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template_submission.docx
  pdf_document: default
csl: ecology.csl
abstract: |
  | **Abstract:** As the average global air temperature on Earth increases, surface temperatures of lakes are also increasing globally (0.34 °C per decade from 1985 to 2009).  The influence of this increased temperature is known to impact lentic ecosystems across local to broad scales.  For example, warming lake temperature is linked to disruptions in trophic linkages, changes in thermal stratification, and cyanobacteria bloom dynamics. Thus understanding broad trends in lake temperature is important to understanding the changing ecology of lakes and the potential human health impacts of these changes.  To help address this, we developed a simple yet robust model of lake photic zone temperature using the 2007 and 2012 US EPA’s National Lakes Assessment data for the conterminous United States. The final model has a mean square error of 2.19 and an adjusted R^2^ of 0.88. The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy. Given the importance of temperature to a lake ecosystem, this model can be a valuable tool for researchers and lake resource managers. Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be estimated based on basic ambient temperature and location information.
  | **Keywords:** Random forest, National Lakes Assessment, Limnology, water temperature, warming lakes
---

```{r setup, include=FALSE, eval=TRUE}
# source(here::here("R/functions.R"), encoding = "UTF-8")
library(captioner)
#opts_chunk$set(echo = TRUE, fig.width=6, fig.height=5)
tab_num <- captioner(prefix = "Table")
fig_num <- captioner(prefix = "Figure")
options(scipen = 2, digits = 5)
```

```{r model, include=FALSE, eval=TRUE, cache=TRUE}
source(here::here("R/model.R"), encoding = "UTF-8")
```
```{r figures, include=FALSE, cache=TRUE}
source(here::here("R/figures.R"), encoding = "UTF-8")
```

@@

For example, warming lakes are linked to a disruption in trophic linkages between phytoplankton and zooplankton [@winder2004climate].  Specifically, Winder and Schindler [-@winder2004climate] report a spring diatom bloom occurring 20 days earlier and a long-term decline in Daphnia populations in Lake Washington, a lake near Seattle, Washington.  Second, warming has been shown to result in changes in thermal stratification across different lake types [@hampton2018change; @oreilly2003climate; @verburg2003ecological; @cohen2016climate; michelutti2016equatorial].  In particular, lakes that typically stratify can exhibit a strengthening of the stratification as well as a reduction in the depth of the thermocline or previously polymictic lakes can begin to stratify as a result of warming lake temperatures. Temperature, in addition to nutrients, is also a key driver to cyanobacteria bloom dynamics [@paerl2012climate].  During periods of higher temperature, cyanobacteria species dominate the phytoplankton community [@lurling2013comparison; @o2012rise; @peperzak2003climate].  As lake temperature increase (typically above 25°C), cyanobacteria have a competitive advantage over phytoplankton and can proliferate quickly [@paerl2008blooms].  Moreover, experimentally enhanced water temperatures yielded significantly increased growth rates of higher toxic Microcystis, but not the non-toxic strains [@davis2009effects].  Thus, our ability to understand and predict toxic cyanobacteria blooms will be deeply dependent on our ability to forecast lake photic zone temperature.  Ultimately, temperature changes will greatly impact every aspect of lake ecology and lake resource management.

Because of the ecological significance, it is not surprising that modelling near-surface lake temperature has been broadly investigated.  Models typically vary in number of lakes studied, complexity of modelling approach, and study interval.  Modelling efforts include efforts to model a single lake over relatively small (e.g. hourly) time intervals [@saeed2016water; @peeters2002modeling; @zhong2016recent].  Additionally, there are numerous studies that model temperature for a small number of lakes while attempting to limit the number of predictor variables [@matuszek1996empirical; @kettle2004empirical, @livingstone1998relationship; @piccolroaz2016prediction; @toffolon2014prediction].  In these efforts, air temperature and lake size are often the only selected predictor variables.  There have also been some efforts aimed at modelling and measuring lake temperature across large spatial extents for large number of lakes [@minns2017factors; @o2015rapid; @wan2017comprehensive].  These past efforts use a wide variety of approaches and have been rather successful at modelling lake photic zone temperature.  One limitation, however, is that the majority of these lakes are large and few studies document or model smaller lakes.  Considering that the number and importance of small lakes (< 1km2) has been historically underestimated, it is important to also predict temperatures in these smaller waterbodies [@downing2006global; @winslow2014lake].

@@

![](../figures/LakePhoticZonev2.jpg)

`r fig_num("LakePhoticZonev2")`

@@

```{r summary, echo=FALSE}
summ <- nla.all %>%
  group_by(year) %>%
  summarize(n = n())
```

@@

![`r fig_num("varsel_fig")`](../figures/varSelFig.jpg)

The final model built with the eight selected variables has a mean-squared error of `r round(mean(RFAll$mse), 2)`  and adjusted R^2^ of `r round(mean(RFAll$rsq), 2)`.  The variables ranked in order of importance were date, average temperature, longitude, 30-day average temperature, elevation, latitude, surface area, and shoreline length (`r fig_num("varImpPlot",caption="Variable importance plot for selected variables.  Shows precent increase in mean square error. Higher values indicate a higher impact on overall model accuracy.", display = "cite")`).  The partial dependency plots illustrate how the predicted photic zone temperature changes over the range of values for all predictor variables (`r fig_num("partPlot",caption="Partial dependence plots for selected variables.", display = "cite")`).

@@

Here we present a simple yet robust model of lake photic zone temperature using the 2007 and 2012 NLA data for the conterminous United States. The final model has a mean square error of `r round(mean(RFAll$mse), 2)` and an adjusted R^2^ of `r round(mean(RFAll$rsq), 2)`. The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy. Given the importance of temperature to a lake ecosystem, especially to cyanobacteria bloom dynamics, this model can be a valuable tool for researchers and lake resource managers. Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be estimated based on basic ambient temperature and location information.

Using only easily obtainable metrics to estimate lake temperature was a primary focus in the creation of this model.  Several studies (e.g. [@piccolroaz2018predictability]) have used similar metrics with good results (RMSE=0.5 °C) to estimate lake temperature but were applied to a single, or a small number of lakes.  Other studies have used more complex models (e.g. General Lake Model (GLM)) to yield good results (RMSE = 1.62 °C for the epilimnion) at varying scales [@bruce2018multi; @hipsey2019general].  However, these models require more complex input variables which are not available for all lakes and may be biased towards large lakes.  Using the NLA data and simple metrics allowed this model to be applicable to any lake within the conterminous United States using minimal input data with a small increase in RMSE.  Additionally, this model is predictive and the model and data will be made publicly available.

The final model included, average ambient air temperature and the average temperature of the prior 30 days. Yet the 3-day and 7-day prior to the sampling date averages were not selected in the final model. It is likely that the three and seven day averages are not providing the model with unique information. Typical the three and seven day averages are fairly close the day-of average. Ambient air temperature does not change drastically over such a short time scale. Even if a big swing in temperature did happen during this short time period, it would be too rare to be a significant factor in the model. Yet the longer term 30-day average does have an impact on lake photic zone temperature. The 30-day average provides us with this information about the temperature intensity leading up to the sample date. Not just how far along into summer; that information would be captured in the date parameter. But more specifically, this measures the long term thermal heating happening at a site.

@@

Despite being one of the most common measurements collected by limnologists, lake temperature datasets that cover long periods of time are very difficult to obtain.  Sharma et al [-@sharma2015global] have compiled summer lake temperature data for 291 lakes for the period 1985-2009.  This may be the largest lake temperature database to date; however, the data are only available to members of their research group and, realistically, the number of lakes included is very small.  One of the reasons we chose to model lake photic zone temperature was to develop a database of lake temperatures for the 48 conterminous United States.  The model we present has proven to be accurate and will allow us to backcast lake temperatures for all the > 300,000 lakes included in NHDplus for the period of time covered by the PRISM climate predictions (1981 to present).  This dataset will allow us to investigate how photic zone temperatures vary both spatially and temporally across the United States.  This database is being developed and, when complete, will be made available as an open source data set.

>>>>

---
title: "**Title:** Lake Photic Zone Temperature across the Conterminous United States"
author:
- ^*1*^ US Environmental Protection Agency, Office of Research and Development,  Atlantic
  Coastal Environmental Sciences Division (ACESD), Narragansett, RI 02882
- ^*2*^ ORISE, Narragansett, RI 02882
- '**Authors:** Kreakie, B. J. ^\*^ ^*1*^,  Shivers, S. ^*2*^, Hollister. J. W. ^*1*^, Milstead,
  W. Bryan. ^*1*^'
- '^\*^  *corresponding author: kreakie.betty@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template_submission.docx
  pdf_document: default
csl: ecology.csl
abstract: |
  | **Abstract:** As the average global air temperature on Earth increases, surface temperatures of lakes are also increasing globally (0.34 °C per decade from 1985 to 2009).  The influence of this increased temperature is known to impact lentic ecosystems across local to broad scales.  For example, warming lake temperature is linked to disruptions in trophic linkages, changes in thermal stratification, and cyanobacteria bloom dynamics. Thus understanding broad trends in lake temperature is important to understanding the changing ecology of lakes and the potential human health impacts of these changes.  To help address this, we developed a simple yet robust model of lake photic zone temperature using the 2007 and 2012 US EPA’s National Lakes Assessment data for the conterminous United States. The final model has a root mean square error of 1.48 °C and an adjusted R^2^ of 0.88. The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy. Given the importance of temperature to a lake ecosystem, this model can be a valuable tool for researchers and lake resource managers. Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be estimated based on basic ambient temperature and location information.
  | **Keywords:** Random forest, National Lakes Assessment, Limnology, water temperature, warming lakes
---

```{r setup, include=FALSE, eval=TRUE}
# source(here::here("R/functions.R"), encoding = "UTF-8")
library(captioner)
library(dplyr)
#opts_chunk$set(echo = TRUE, fig.width=6, fig.height=5)
tab_num <- captioner(prefix = "Table")
fig_num <- captioner(prefix = "Figure")
options(scipen = 2, digits = 5)
```

```{r model, include=FALSE, eval=TRUE}
source(here::here("R/model.R"), encoding = "UTF-8")
```
```{r figures, include=FALSE}
source(here::here("R/figures.R"), encoding = "UTF-8")
```

@@

For example, warming lakes are linked to a disruption in trophic linkages between phytoplankton and zooplankton [@winder2004climate].  Specifically, Winder and Schindler [-@winder2004climate] report a spring diatom bloom occurring 20 days earlier and a long-term decline in Daphnia populations in Lake Washington, a lake near Seattle, Washington.  Second, warming has been shown to result in changes in thermal stratification across different lake types [@hampton2018change; @oreilly2003climate; @verburg2003ecological; @cohen2016climate; @michelutti2016equatorial].  In particular, lakes that typically stratify can exhibit a strengthening of the stratification as well as a reduction in the depth of the thermocline or previously polymictic lakes can begin to stratify as a result of warming lake temperatures. Temperature, in addition to nutrients, is also a key driver to cyanobacteria bloom dynamics [@paerl2012climate].  During periods of higher temperature, cyanobacteria species dominate the phytoplankton community [@lurling2013comparison; @o2012rise; @peperzak2003climate].  As lake temperature increase (typically above 25°C), cyanobacteria have a competitive advantage over phytoplankton and can proliferate quickly [@paerl2008blooms].  Moreover, experimentally enhanced water temperatures yielded significantly increased growth rates of higher toxic Microcystis, but not the non-toxic strains [@davis2009effects].  Thus, our ability to understand and predict toxic cyanobacteria blooms will be deeply dependent on our ability to forecast lake photic zone temperature.  Ultimately, temperature changes will greatly impact every aspect of lake ecology and lake resource management.

Because of the ecological significance, it is not surprising that modelling near-surface lake temperature has been broadly investigated.  Models typically vary in number of lakes studied, complexity of modelling approach, and study interval.  Modelling efforts include efforts to model a single lake over relatively small (e.g. hourly) time intervals [@saeed2016water; @peeters2002modeling; @zhong2016recent].  Additionally, there are numerous studies that model temperature for a small number of lakes while attempting to limit the number of predictor variables [@matuszek1996empirical; @kettle2004empirical; @livingstone1998relationship; @piccolroaz2016prediction; @toffolon2014prediction].  In these efforts, air temperature and lake size are often the only selected predictor variables.  There have also been some efforts aimed at modelling and measuring lake temperature across large spatial extents for large number of lakes [@minns2017factors; @o2015rapid; @wan2017comprehensive].  These past efforts use a wide variety of approaches and have been rather successful at modelling lake photic zone temperature.  One limitation, however, is that the majority of these lakes are large and few studies document or model smaller lakes.  Considering that the number and importance of small lakes (< 1km2) has been historically underestimated, it is important to also predict temperatures in these smaller waterbodies [@downing2006global; @winslow2014lake].

@@

@@

```{r summary, include=FALSE, message=FALSE}
summ <- nla.all %>%
  group_by(year) %>%
  summarize(n = n())

extreme <- nla_select %>%
  filter(tmean_2m < 15 | tmean_2m > 32) %>%
  summarize(perc = nrow(.)/nrow(nla_select) *100)
```

@@

<!-- Check y-axis label -->
![`r fig_num("varsel_fig")`](../figures/varSelFig.jpg)

The final model built with the eight selected variables has a root mean-squared error of `r round(sqrt(mean(RFAll$mse)), 2)`  and adjusted R^2^ of `r round(mean(RFAll$rsq), 2)`. The model performs well across a wide range of temperatures (`r fig_num("obs_v_pred", caption = "Predicted versus measured photic zone temperature", display = "cite")`); however, at the higher and lower temperature (i.e. less than 15 °C and greater than 32 °C) the model does not perform as well but lakes with growing season temperatures at these extremes were rare and represented only `r round(extreme$perc, 2)` of lakes sampled in the 2007 and 2012 NLA.

![`r fig_num("obs_v_pred")`](../figures/obs_v_pred.jpg)

The variables ranked in order of importance were date, average temperature, longitude, 30-day average temperature, elevation, latitude, surface area, and shoreline length (`r fig_num("varImpPlot",caption="Variable importance plot for selected variables.  Shows precent increase in mean square error. Higher values indicate a higher impact on overall model accuracy.", display = "cite")`).  The partial dependency plots illustrate how the predicted photic zone temperature changes over the range of values for all predictor variables (`r fig_num("partPlot",caption="Partial dependence plots for selected variables.", display = "cite")`).

@@

Here we present a simple yet robust model of lake photic zone temperature using the 2007 and 2012 NLA data for the conterminous United States. The final model has a root mean square error of `r round(sqrt(mean(RFAll$mse)), 2)` and an adjusted R^2^ of `r round(mean(RFAll$rsq), 2)`. The sampling date, that day’s average ambient air temperature, data obtained from the PRISM Climate Group, and longitude are the most important variables impacting the final model’s accuracy. Given the importance of temperature to a lake ecosystem, especially to cyanobacteria bloom dynamics, this model can be a valuable tool for researchers and lake resource managers. Daily predicted lake photic zone temperature for all lakes in the conterminous US can now be estimated based on basic ambient temperature and location information.

Using only easily obtainable metrics to estimate lake temperature was a primary focus in the creation of this model.  Several studies (e.g. [@piccolroaz2018predictability]) have used similar metrics with good results (RMSE = 0.5 °C) to estimate lake temperature but were applied to a single, or a small number of lakes.  Other studies have used more complex models (e.g. General Lake Model (GLM)) to yield good results (RMSE = 1.62 °C for the epilimnion) at varying scales [@bruce2018multi; @hipsey2019general].  However, these models require more complex input variables which are not available for all lakes and may be biased towards large lakes.  Using the NLA data and simple metrics allowed this model to be applicable to any lake within the conterminous United States using minimal input data with a comparable RMSE.

The final model included, average ambient air temperature and the average temperature of the prior 30 days. Yet the 3-day and 7-day prior to the sampling date averages were not selected in the final model. It is likely that the three and seven day averages are not providing the model with unique information as the three and seven day averages are likely to be similar to the day-of average. Ambient air temperature does not change drastically over such a short time scale [CITATION??]. Even if a big swing in temperature did happen during this short time period, it would be too rare to be a significant factor in the model.<!--JWH: We can quantify this with the data that we have.  Should probably do this.  Less arm wavey --> Yet the 30-day average does have an impact on lake photic zone temperature as it provides information about the longer term temperature intensity leading up to the sample date. Not just how far along into summer; that information would be captured in the date parameter. But more specifically, this measures the long term thermal heating happening at a site.

@@

Despite being one of the most common measurements collected by limnologists, lake temperature datasets that cover long periods of time are very difficult to obtain.  Sharma et al [-@sharma2015global] have compiled summer lake temperature data for 291 lakes for the period 1985-2009.  This may be the largest lake temperature database to date; however, the data are only available to members of their research group and, realistically, the number of lakes included is very small. <!--These data are open and published - accessible through the "laketemps" package.-->  One of the reasons we chose to model lake photic zone temperature was to develop a database of lake temperatures for the 48 conterminous United States.  The model we present has proven to be accurate and will allow us to backcast lake temperatures for all the > 300,000 lakes included in NHDplus for the period of time covered by the PRISM climate predictions (1981 to present).  This dataset will allow us to investigate how photic zone temperatures vary both spatially and temporally across the United States.  This database is being developed and, when complete, will be made available as an open source data set.
